Polynomial Time and Dependent Types

### 2023-07-11 paper/paper.tex
[ paper ] Complete paper?

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Polynomial Time and Full Dependent Types}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms. We prove the soundness of the systems
  using a realisability technique due to Dal Lago and Hofmann.

@@

In this paper, we propose a method for extending dependent type theory
with a means for constraining the intrinsic computational complexity
of programs written in the theory. We concentrate on linear type
systems that soundly and completely capture polynomial time
computation, the commonly used standard for feasible resource usage,
and extend these systems to dependent types. The additional
expressivity of dependent types allows us use these characterisations
of polytime to further functionally characterise

We use techniques from \emph{Implicit Computational Complexity}
theory, which seeks to provide intrinsic characterisations of
complexity classes in terms of logical systems or programming
languages. We review the techniques that we use in
\autoref{sec:linear-intro}. To adapt these systems to dependent types,
we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt,mcbride16}, a combination of linear and dependent
types. We review QTT in \autoref{sec:qtt}.

Our long-term goal is to combine the extensional reasoning of type
theory with intensional reasoning about the resources intrinsically
consumed by programs. This paper is a first step along this path,
which we hope will lead both to practical systems for reasoning about
programs' resource usage, and to theoretical use as a form of
\emph{synthetic computational complexity theory}. We discuss these
possibilities further in \autoref{sec:conclusion}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Affine Linear $\lambda$-Calculus and Polytime}
\label{sec:linear-intro}

Not long after Girard introduced Linear Logic \cite{girard87}, it was
observed that its resource sensitivity could be turned to describing
computational complexity classes by purely logical means. Typically, a
logical system is described for which the process of reducing a proof
to a normal form (often by cut elimination) is guaranteed to always be
accomplished within a certain complexity bound. Moreover, the system
is usually proven to be complete for the relevant complexity class, by
constructing a simulation of some known representation. Such systems
that characterise polytime include Bounded Linear Logic (BLL), which
uses explicit polynomials in the formulas \cite{bll} and Soft Affine
Logic (SAL) \cite{lafont}, which does not explicitly represent
time information in formulas, but uses a restricted form of Linear
Logic's $\oc$ modality instead. Light Affine Logic (LAL) \cite{lal} is
another ``counting-free'' system for polytime.

Viewing logical systems though the lens of the Curry-Howard
correspondence, the idea arises that one could define functional
programming languages that characterise complexity classes such as
polytime. SAL has been transformed into a programming language for
polytime by \cite{gaboardi}. Hofmann proposed a new programming
language, Linear Functional Programming Language (LFPL), that uses a
novel ``payment'' system to track iteration. In this section, we
review the use of linear types to capture polytime by presenting two
systems, one based on ideas from SALL and the second more explicitly
based on LFPL. With a view to adapting these systems to dependent
types in \autoref{sec:qtt}, we take an approach slightly different to
much of the polytime linear logic literature. We emphasise the use of
explicit datatypes and eliminators, rather than using impredicative
encodings via universal types. Thus, we are closer to Hofmann's
original LFPL (though not later presentations of it
\cite{dallago-hofmann}) than BLL, SALL, or LAL.

@@

It is not too difficult to see that reduction of linear
$\lambda$-terms always takes a number of steps linearly proprotional
to the size of the term. This is because every $\beta$-redex
substitutes into exactly one variable, reducing the size of the term
by one each time.

We can increase the expressivity, but not the computational
complexity, of the system by adding datatypes that do not allow
iteration. These can be used for representation, but not for driving
computation. We include the rules here, to show how linearity must be
preserved in these rules, and to foreshadow their dependently typed
counterparts in \autoref{sec:qtt-noniter}. The first type is the
booleans, which are anyway non-recursive and so wouldn't allow
iteration anyway:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cTrue, \cFalse : \BoolTy}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \ListTy(A) \\
    \Gamma_2 \vdash N_1 : B \\
    \Gamma_2, h : A, t : \ListTy(A) \vdash N_2 : B}
  {\Gamma_1, \Gamma_2 \vdash \Match\,M\,\{\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\} : B}
\end{mathpar}
Hence, we can construct lists arbitrarily, but we can only do case
analysis on them. If we wish to explore a list to a arbitrary depth it
must be driven by an iterable type of the form that we will see in the
next two sections.

With booleans and lists, we can construct several other useful
types. For example, to simulate Turing machines, one can construct a
$\mathrm{Tape}$ type as a Zipper \cite{huet}
$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
representing a position on the tape with the items before, under, and
after the head.

@@

Polynomial time is usually seen as a proxy for ``feasible''
computation. On the face of it, there does not seem to be any
particular reason why polynomials have anything to do with
feasibility. However, one can arrive at the definition of polynomial
time in three steps, by assuming that (i) iterating over the whole
input is feasible; (ii) if two computations are feasible, then so is
their composition; and (iii) performing a feasible computation for
every element of the input is also feasible. It is the last point that
allows complexities of arbitrary polynomial degree to be constructed
(we will see this in action in the completeness construction below and
soundness proofs in \autoref{sec:iterator-soundness}).

@@

A more flexible system was proposed by \cite{hofmann99}. The
\emph{Linear Functional Programming Language} (LFPL) takes a different
approach to the \emph{Cons-free} system. Instead of completely
prohibiting construction of new data, constructions is allowed if it
is paid for by values of new type $\Diamond$ (``diamonds''):
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M : \Diamond}
  {\Gamma \vdash \conZero(M) : \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
Thus, to construct a $\conZero$, we must have a $\Diamond$ to pay for
it, and likewise, to construct a $\conSucc$ we must have a diamond. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an addition binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to Aelhig and
Schwictenberg \cite{syntactic-lfpl}, illustrates how this leads to all
polytime computations. As above, we assume a polynomial $p(n)$ and a
step function $f : \mathit{St} \multimap \mathit{St}$ that needs to be
iterated $p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather the iterator $I_k$
performs $\binom{n}{k}$ iterations. As observed by Aelhig and
Schwictenberg, this is sufficient because the binomials form a basis
for the vector space of all polynomials.

@@

We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
capturing polytime by means of linear typing and restricted
iteration. We now look to extend these systems to include dependent
types.

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. We now review the definition of QTT and describe
how we have adapted it to the polytime systems we saw in the previous
section.

@@

QTT is a system that resolves this conflict by recording usage of
variables with annotations from a semiring. It sits in the general
area of systems that use semiring annotatons to measure resource usage
\cite{BLAH}. The key feature of QTT, originally due to
\cite{mcbride16}, is that usage of variables in types counts for
$0$-usage in terms of the semiring used. This allows us to use normal
type theory as a specification language, while also enjoying the
benefits of linear typing for programs. The term typing judgement of
QTT has the following form:
\begin{displaymath}
  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
\end{displaymath}
where the annotations $\rho_i$ are all from the semiring being
used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
we are in the erased ($\sigma = 0$) fragment, where all the normal
rules of type theory apply, or the in the non-erased (``present'',
$\sigma = 1$) fragment, where a restricted typing discipline applies.
As we shall see below in the cases of $\Sigma$-types, iterable types,
and LFPL's $\Diamond$ type, the separation of QTT into these two
fragments allows an expressive combination of reasoning using full
type theory with the benefits of linear typing.

In the remainder of this sub-section, we describe the core of QTT. As
well as the term typing judgement given above, QTT also has judgements
for well-formed contexts ($\Gamma~\isctxt$) and types
($\Gamma \vdash T~\istype$), and definition equality of types and
terms ($\Gamma \vdash S \equiv T~\istype$ and
$\Gamma \vdash M \equiv N \stackrel\sigma: S$). It is an invariant of
the system that types are always well-formed in a context with all
annotations $0$, i.e., $\Gamma \vdash S~\istype$ implies
$0\Gamma = \Gamma$. An important admissible rule of the system, along
with weakening and substitution, is that of $0$-ing:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M \stackrel1: S}
  {0\Gamma \vdash M \stackrel0: S}
\end{displaymath}
This rule allows us to take any term $M$ in the $\sigma = 1$ fragment
and treat it as if it were in the $\sigma = 0$ fragment, and hence use
it for specification purposes in types. As we add novel rules to QTT
in the following sections, we will be careful to maintain the
admissibility of this rule.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$.

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about
implementability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}
QTT also supports a unit type $I$ with constructor $*$ and
pattern-matching elimination. See \cite{atkey18qtt} for details.

@@

  \inferrule*
  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
\end{mathpar}
The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
as the only proof of equality \cite{hoffmann99dependent}. Note that
equality reflection only targets the $\sigma = 0$ fragment, we cannot
use equality reflection to convey realisability information.

\subsubsection{Universe} QTT can also be equipped in universe types
$\mathsf{U}$, just as in standard type theory. We refer to the rules
presented in \cite{atkey18qtt}.

\subsubsection{Datatypes} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:linear-intro}, we add two kinds of datatype to QTT. First,
\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
construction and case analysis, but no recursion. In
\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

@@

\subsubsection{Booleans} The boolean type for QTT was described in
\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
offer no possibility for iteration, but it is useful to see how the
QTT versions of the rules extend the simply typed rules from
\autoref{sec:linear-case-analysis} before moving to more complex
types.
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \BoolTy~\istype}

@@

The reader is invited to compare this dependently typed rule with the
simply typed linear version in \autoref{sec:cons-free-intro}. Removing
the $0$-annotated parts of the rule, and the type dependency, yield
the exact same rule. Conversely, when $\sigma = 0$, this rule is
identical (up to $0$-annotations) to the usual dependently typed
recursor for natural numbers, and so we can use it in the types to
prove properties of programs just as we do in normal type theory. We
will see in \autoref{sec:cons-free-sound} that this rule is realisable
by polynomial-time amortised computation, and so is sound for
polynomial time.

\subsection{LFPL-style Diamonds and Natural Numbers}
\label{sec:lfpl-qtt}

@@

% \begin{enumerate}
% \item Non-deterministic polytime via a bit oracle -- needs a statement
%   of success
% \item Previous could also be interpreted as probabilistic polytime
%   (access to coin flips); can we formalise some kind of cryptographic
%   property this way? Have a look at hash functions in the
%   Authenticated Data Structures paper.
% \item Insertion sort, quicksort as polytime functions.
% \item A type of polytime functions; via a reflection operator
% \item Programming with explicit resource annotations, including
%   reasoning about the resource requirements directly.
% \end{enumerate}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Polynomial Time and Dependent Types}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. We explore two
  systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types via
  Quantitative Type Theory, allowing for arbitrary computation in
  types alongside guaranteed polynomial time computation in terms. We
  prove the soundness of the systems using a realisability technique
  due to Dal Lago and Hofmann.

@@

In this paper, we propose a method for extending dependent type theory
with a means for constraining the intrinsic computational complexity
of programs written in the theory. We concentrate on linear type
systems that soundly and completely capture polynomial time
computation, the commonly used standard for feasible resource usage,
and extend these systems to dependent types. The additional
expressivity of dependent types allows us use these characterisations
of polytime to further functionally characterise the classes of
non-deterministic and probabilistic polynomial time.

We use techniques from \emph{Implicit Computational Complexity}
theory, which provides intrinsic characterisations of complexity
classes in terms of logical systems or programming languages. We
review the techniques that we use in \autoref{sec:linear-intro}. To
adapt these systems to dependent types, we use \emph{Quantitative Type
  Theory} (QTT) \cite{atkey18qtt,mcbride16}, a combination of linear
and dependent types. We review QTT in \autoref{sec:qtt}.

Our long-term goal is to combine the extensional reasoning of Type
Theory with intensional reasoning about the resources intrinsically
consumed by programs. This paper is a first step along this path,
which we hope will lead both to practical systems for reasoning about
programs' resource usage, and to theoretical use as a form of
\emph{synthetic computational complexity theory}. We discuss these
possibilities further in \autoref{sec:conclusion}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Affine Linear Typing and Polytime}
\label{sec:linear-intro}

Not long after Girard introduced Linear Logic \cite{girard87}, it was
observed that its resource sensitivity could be turned to describing
computational complexity classes by purely logical means. Typically, a
logical system is described for which the process of reducing a proof
to a normal form (often by cut elimination) is guaranteed to always be
accomplished within a certain complexity bound. Moreover, the system
is usually proven to be complete for the relevant complexity class, by
constructing a simulation of some known representation. Such systems
that characterise polytime include Bounded Linear Logic (BLL), which
uses explicit polynomials in the formulas \cite{bll92} and Soft Affine
Logic (SAL) \cite{Lafont04}, which does not explicitly represent time
information in formulas, but uses a restricted form of Linear Logic's
$\oc$ modality instead. Light Linear Logic (LLL) \cite{lll98} is
another ``counting-free'' system for polytime.

Viewing logical systems though the lens of the Curry-Howard
correspondence, the idea arises that one could define functional
programming languages that characterise complexity classes such as
polytime. SAL has been transformed into a programming language by
\citet{BaillotM04}, and likewise for LLL by
\citet{BaillotGM10}. \citet{hofmann99lfpl} proposed a new programming
language, Linear Functional Programming Language (LFPL), that uses a
novel ``payment'' system to track iteration. In this section, we
review the use of linear types to capture polytime by presenting two
systems, one based on ideas from SAL and the second more explicitly
based on LFPL. With a view to extending to dependent types in
\autoref{sec:qtt}, we take an approach slightly different to much of
the polytime linear logic literature. We emphasise the use of explicit
datatypes and eliminators, rather than using impredicative encodings
via universal types. We are closer to Hofmann's original LFPL (though
not a later presentation of it by \citet{dallago11realisability}) than
BLL, SAL or LLL.

@@

It is not too difficult to see that reduction of linear
$\lambda$-terms always takes a number of steps linearly proprotional
to the size of the term. This is because every $\beta$-redex
substitutes into at most one variable, reducing the size of the term
by one each time.

We can increase the expressivity, but not the computational
complexity, of the system by adding datatypes that do not allow
iteration. These can be used for representation, but not for driving
computation. We include the rules here, to show how linearity must be
preserved in these rules, and to foreshadow their dependently typed
counterparts in \autoref{sec:noniter-qtt}. The first type is the
booleans, which are anyway non-recursive and so would not allow
iteration anyway:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cTrue, \cFalse : \BoolTy}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \ListTy(A) \\
    \Gamma_2 \vdash N_1 : B \\
    \Gamma_2, h : A, t : \ListTy(A) \vdash N_2 : B}
  {\Gamma_1, \Gamma_2 \vdash \Match\,M\,\{\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\} : B}
\end{mathpar}
We can construct lists arbitrarily but only do case analysis on
them. If we wish to explore a list to a arbitrary depth it must be
driven by a type we can iterate over.

With booleans and lists, we can construct several other useful
types. For example, to simulate Turing machines, one can construct a
$\mathrm{Tape}$ type as a Zipper (\citet{Huet97})
$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
representing a position on the tape with the items before, under, and
after the head.

@@

Polynomial time is usually seen as a proxy for ``feasible''
computation. On the face of it, there does not seem to be any
particular reason why polynomials have anything to do with
feasibility. However, one can arrive at the definition of polynomial
time in three steps, by assuming that (i) iterating over the whole
input is feasible; (ii) if two computations are feasible, then so is
their composition; and (iii) performing a feasible computation for
every element of the input is also feasible. It is the last point that
allows complexities of arbitrary polynomial degree to be constructed
(we will see this in action in the completeness construction below and
soundness proofs in \autoref{sec:realising-iteration}).

@@

A more flexible system was proposed by \citet{hofmann99lfpl}. Instead
of completely prohibiting construction of data, the \emph{Linear
  Functional Programming Language} (LFPL) allows construction if it is
paid for by values of type $\Diamond$ (``diamonds''):
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M : \Diamond}
  {\Gamma \vdash \conZero(M) : \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather the iterator $I_k$
performs $\binom{n}{k}$ iterations. As observed by Aelhig and
Schwictenberg, this is sufficient because the binomials form a basis
for the vector space of all polynomials.

@@

We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
capturing polytime by means of linear typing and restricted
iteration. We now look to extend these systems to include dependent
types by building upon \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt,mcbride16}. This section reviews QTT and describes
how we have adapted it to the polytime systems we saw in the previous
section.

@@

QTT is a system that resolves this conflict by recording usage of
variables with annotations from a semiring. It sits in the general
area of systems that use semiring annotatons to measure resource usage
\cite{BrunelGMZ14,GhicaS14,OrchardLE19}. The key feature of QTT, an
insight owing to \citet{mcbride16}, is that usage of variables in
types counts for $0$-usage in terms of the semiring used. This allows
us to use normal type theory as a specification language, while also
enjoying the benefits of linear typing for programs. The term typing
judgement of QTT has the following form:
\begin{displaymath}
  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
\end{displaymath}
where the annotations $\rho_i$ are all from the semiring being
used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
we are in the erased ($\sigma = 0$) fragment, where all the normal
rules of type theory apply, or the in the non-erased (``present'',
``realisable'', $\sigma = 1$) fragment, where a restricted typing
discipline applies.  As we shall see below in the cases of
$\Sigma$-types, iterable types, and LFPL's $\Diamond$ type, the
separation of QTT into these two fragments allows an expressive
combination of reasoning using full type theory with the benefits of
linear typing.

In the remainder of this sub-section, we describe the core of QTT. As
well as the term typing judgement given above, QTT also has judgements
for well-formed contexts ($\Gamma~\isctxt$) and types
($\Gamma \vdash T~\istype$), and definitional equality of types and
terms ($\Gamma \vdash S \equiv T~\istype$ and
$\Gamma \vdash M \equiv N \stackrel\sigma: S$). It is an invariant of
the system that types are always well-formed in a context with all
annotations $0$, i.e., $\Gamma \vdash S~\istype$ implies
$0\Gamma = \Gamma$. An important admissible rule of the system, along
with weakening and substitution, is that of $0$-ing:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M \stackrel1: S}
  {0\Gamma \vdash M \stackrel0: S}
\end{displaymath}
This rule allows us to take any term $M$ in the $\sigma = 1$ fragment
and treat it as if it were in the $\sigma = 0$ fragment, and hence use
it for specification purposes in types. As we add novel rules to QTT
in the following sections, we will be careful to maintain the
admissibility of this rule.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$, where $x$ does not appear in $T$.

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about polytime
realisability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
$\Sigma$-types are unrestricted in the $\sigma = 0$ fragment, and we
can use them as normal for type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}
QTT also supports a unit type $I$ with constructor $*$ and
pattern-matching \cite{atkey18qtt}.

@@

  \inferrule*
  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
\end{mathpar}
The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
as the only proof of equality \cite{hofmann97syntax}. Note that
equality reflection only targets the $\sigma = 0$ fragment, we cannot
use equality reflection to convey realisability information.

\subsubsection{Universe} QTT has universe types $\mathsf{U}$, as in
standard type theory \cite{atkey18qtt}.

\subsubsection{Datatypes} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:linear-intro}, we add two kinds of datatype to
QTT. First, in \autoref{sec:noniter-qtt}, we add non-iterable
datatypes that allow construction and case analysis, but no
recursion. In \autoref{sec:cons-free-qtt} we describe how to extend
QTT to be a dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

@@

\subsubsection{Booleans} The boolean type for QTT was described in
\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
offer no possibility for iteration, but it is useful to see how the
QTT rules extend the simply typed rules from
\autoref{sec:linear-case-analysis} before moving to more complex
types.
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \BoolTy~\istype}

@@

The reader is invited to compare this dependently typed rule with the
simply typed linear version in \autoref{sec:cons-free-intro}. Removing
the $0$-annotated parts of the rule, and the type dependency, yield
the exact same rule. Conversely, when $\sigma = 0$, this rule is
identical (up to $0$-annotations) to the usual dependently typed
recursor for natural numbers, and so we can use it in the types to
prove properties of programs just as we do in normal type theory. We
will see in \autoref{sec:consfree-sound} that this rule is realisable
by polynomial-time computation, and so is sound for polynomial time.

\subsection{LFPL-style Diamonds, Natural Numbers, and a Recursor that Gives Back}
\label{sec:lfpl-qtt}

@@
